A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-08-08 code/supplements.Rmd
Add Figure 4 forest heatmap with figure legend. Update text to reflect figure 4. Adjust manuscript.rmd to work with docx

```{r}
library(data.table)
library(ggplot2)
library(tidyverse)
library(kableExtra)
```

### Supplementary Figure S1
```{r supplementary-figure-s1, out.width = '90%', fig.align='center'}
umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data

@@

```{r supplementary-figure-s2-data-generation, out.width = '90%', fig.align = "center"}
umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet

@@

<div style="text-align: right"> Supplementary Figure S3 Data </div>
```{r supplementary-figure-s3-data-generation, eval = FALSE}
load("../data/gdc_tumor_coding_complete.Rdata")
load("../data/gene_data.Rda")

@@

```{r supplementary-figure-s3, out.width='77%', fig.align = "center"}
load("../data/selected_mut.Rda")
umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet

@@

<div style="text-align: right"> Supplementary Figure S4 Data </div>
```{r supplementary-figure-s4-data-generation, eval = FALSE}
load("../data/gdc_reactome_path_analysis.Rda") ## load data
umap_cancertype <- fread("../data/umap_3d_coors.tsv") 
n_path <- 377 ## how many pathways do we want to look at starting from the most to least mutated pathway

@@

```{r supplementary-figure-s4, message = FALSE, echo = FALSE, out.width = '90%', fig.align= 'center'}
load("../data/fig_s4_ggplot.Rda")

@@

<div style="text-align: right"> Supplementary Figure ST1 Data </div>
```{r supplemental-table-st1-data-generation, eval = FALSE}
load("../data/gdc_tumor_coding_complete.Rdata")
load("../data/gdc_reactome_path_analysis.Rda")
load("../data/gene2_level2_lookup.Rda")
load("../data/pathname2level2.Rda")
load("../data/gene_data.Rda")
load("../data/gdc_gene_exp_adj.Rda")

@@

```{r supplemental-table-st1}
load("../data/table_st1.Rda")

table_st1 %>% dplyr::select(-Pathway) %>% knitr::kable(format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
  column_spec(1, bold = T, color = "black")
```

![**Supplementary Table ST1. Frequency of pathway disruption and top genes within pathways.** _A complete list of pathway disruptions with percent mutatied samples and top genes_]

>>>>

### Supplementary Figure S1

```{r supplementary-figure-s1, fig.width = 9, fig.height= 8, fig.align='center'}
#out.width = '90%',
umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data

@@

```{r supplementary-figure-s2-data-generation, fig.width = 9, fig.height= 8, fig.align = "center"}
umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet

@@

```{r supplementary-figure-s3-data-generation, eval = FALSE}
load("../data/gdc_tumor_coding_complete.Rdata")
load("../data/gene_data.Rda")

@@

```{r supplementary-figure-s3, fig.width = 7, fig.height= 5, fig.align = "center"}
# out.width = '77%'
load("../data/selected_mut.Rda")
umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet

@@

```{r supplementary-figure-s4-data-generation, eval = FALSE}
load("../data/gdc_reactome_path_analysis.Rda") ## load data
umap_cancertype <- fread("../data/umap_3d_coors.tsv") 
n_path <- 377 ## how many pathways do we want to look at starting from the most to least mutated pathway

@@

```{r supplementary-figure-s4, message = FALSE, echo = FALSE, out.width = '90%', fig.align= 'center', eval = FALSE}
load("../data/fig_s4_ggplot.Rda")

@@

```{r supplemental-table-st1-data-generation, eval = FALSE}
load("../data/gdc_tumor_coding_complete.Rdata")
load("../data/gdc_reactome_path_analysis.Rda")
load("../data/gene2_level2_lookup.Rda")
load("../data/pathname2level2.Rda")
load("../data/gene_data.Rda")
load("../data/gdc_gene_exp_adj.Rda")

@@

```{r supplemental-table-st1, eval = FALSE}
load("../data/table_st1.Rda")
## for html only
# table_st1 %>% dplyr::select(-Pathway) %>% knitr::kable() %>%
#   kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
#   column_spec(1, bold = T, color = "black")

## for word doc
# table_st1 %>% flextable() %>% 
#   set_caption(caption = "") %>%
#   align(part = "all") %>% 
#   width(., width = c(4,0.5,0.5,0.5,0.5)) %>% 
#   fontsize(size = 8, part = "body")

write.xlsx(table_st1, file = "../data/table_st1.xlsx", row.names = FALSE)
```

```{r enrich-pathway-in-clust}
## docx friendly
load("../data/moderate_enrich_prob.Rda")

st1 <- moderate_enrich_prob %>% filter(difference >= 0.30 & signif <= 2500) %>% ## filter effect size >= 0.20 and enrichment at a 95% CI
  dplyr::select(path, clust, signif, difference) %>% 'colnames<-' (c("Pathway", "Cluster", "Signif", "Enrichment_Score")) ## select relevant column and rename col

st1$log10pvalue <- round(-log10((st1$Signif + 1) / 1E5), digits = 2) ## Getting p value ## adding 1 for pseudo count.. 1E5 is the n simulations performed
st1$Enrichment_Score <- round(st1$Enrichment_Score, digits = 2)
st1$Signif <- NULL ## we don't need this anymore

st1 %>% flextable() %>% 
  set_caption(caption = "") %>%
  align(part = "all") %>% 
  width(., width = c(4,0.75,1,1)) %>% 
  fontsize(size = 10, part = "body")
```

### 2020-08-08 text/manuscript.Rmd
Add Figure 4 forest heatmap with figure legend. Update text to reflect figure 4. Adjust manuscript.rmd to work with docx

---
title: "A molecular taxonomy of tumors independent of tissue-of-origin"
author: "Peter T. Nguyen$^{1}$, Simon G. Coetzee$^{1}$, Daniel L. Lakeland$^{2}$, and Dennis J. Hazelett$^{1,3}$"
output: 
  bookdown::html_document2:
    fig_captions: TRUE
    number_sections: FALSE
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
    code_folding: hide
    theme: flatly
    smart: true
bibliography: references.bib
editor_options: 
  chunk_output_type: console
---

@@

```{r echo=FALSE, eval=FALSE}
  bookdown::pdf_book:
    number_sections: FALSE
    toc: FALSE
    keep_tex: yes
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages(c(
  library(rstan),
  library(foreach),
  library(sqldf),
  library(tidyverse),
  library(ggthemes),
  library(data.table),
  library(biomaRt),
  library(dplyr),
  library(GenomicRanges),
  library(SummarizedExperiment),
  library(reactome.db),
  library(plyr),
  library(cluster),
  library(pheatmap),
  library(clusterProfiler),
  library(org.Hs.eg.db),
  library(dbscan),
  library(class),
  library(TCGAbiolinks),
  library(patchwork),
  library(ggplotify),
  library(ComplexHeatmap),
  library(knitr),
  library(kableExtra),
  library(UpSetR),
  library(cowplot),
  library(doRNG),
  library(uwot),
  library(janitor),
  library(FactoMineR),
  library(seriation),
  library(rtracklayer),
  library(poibin)
))
```

@@

```{r child='../code/figure2.Rmd'}
```

@@

### Gene-Pathway redundancy contributes to cluster identity

@@

```{r child='../code/figure4_function.Rmd'}
```

@@

Third, we checked the mutation classes using MAFtools [@maftools] and found the overwhelming
majority (x out of y) are missense _vs._ various other deleterious classes (see
**figure 3D**), suggesting that, similar to other oncogenic driver classes such
as KRAS and PIK3C, proteasomal subunit mutations may result in specific
dysregulation of the proteasomal complex activities that target cancer-promoting
pathways. Taken together, our data suggest that mutations in proteasomal
subunits are more widespread in cancer than previously appreciated and may
represent a targetable category of tumors.

@@

<mark>As discussed previously, prior efforts to extract common signatures from
pan-cancer datasets met with difficulty in distinguishing tumor samples from
their tissue-specific -omics data signatures. Given our pathway-disruption based
classification, this raises the question, are tumor phenotypes driven entirely
by common driver genes, or by "silent" tissue-specific effectors ( _i.e._ too
few samples to detect above statistical significance thresholds), or a
combination of both? To answer this question, we ranked the top cluster-specific
pathways for each cluster, and then tabulated the most frequently mutated genes
for each pathway as a function of cancer type (**Table 1** and **Supplementary tables ST2-ST13**). 
We find that in most clusters the top pathways are
determined by mutations in a handful of driver genes, TP53, PIK3CA, KRAS etc.
However in many cancers we find divergence outside these top driver mutations,
wherein etc etc</mark>

```{r child='../code/table1.Rmd'}
```

@@

Therefore, we calculated cluster-specific enrichment in metastatic tumors and
found a total of 27 enriched pathways across all clusters (**table 2**). A number of
enrichments were found in multiple clusters represented by pathways that were
already shown to be enriched in a neighboring cluster. For example,
“Signaling by PTK6” is enriched in non-metastatic samples of cluster 8 (see 
**supplementary figure S4**) but not in 7 and 9. This pathway is enriched in
metastatic tumors of clusters 7 and 9 (p < 10-3, **table 2**). This is also true of 
“Erythropoietin activates RAS”, which is enriched in cluster 8 (**supplementary figure
S4**) and also in metastatic tumors of clusters 7 and 9. <strike>Cluster 1, one of two clusters 
for which we could not identify any distinguishing pathways (cluster 8 being the other), 
had enrichment in two pathways that were otherwise specific to non-metastatic tumors of other
clusters. These include “Cellular response to hypoxia” (enriched in clusters 6 &
7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7).</strike> Cluster 4 metastatic
samples was enriched for “Fc epsilon receptor (FCERI) signaling”, a key neutrophil pathways, 
which are specific to clusters 2, 7 and 8 non-metastatic tumors. Thus, pathway lesions 
specific to metastases tend toward complementary mutations that are otherwise specific to
non-metastatic tumors of other clusters.

```{r child='../code/table2.Rmd'}
```
<mark>Statement does not apply anymore. The highest # of enriched path is 3. Not many are in 
metastasis. Only 2. “Fc epsilon receptor (FCERI) signaling” and "Signaling by SCF-KIT" </mark>
<strike>There were also a number of pathways that were enriched with lower specificity
(among 4 or more clusters) which were found to be enriched in metastases. We
identified 18 such pathways of which 12 had significantly higher mutation rates
in cluster-specific metastases, including “G1/S DNA Damage Checkpoints”,
“Deubiquitination”, “Sumoylation”, “M Phase”, “RHO GTPase Effectors”, “TCF
dependent signaling in response to WNT”, “Class I MHC mediated antigen
processing & presentation”, “Chaperonin-mediated protein folding”, “Intrinsic
Pathway for Apoptosis”, “Cellular Senescence”, “DNA Double-Strand Break
Response”, and “Pre-NOTCH Expression and Processing” (**Table 2**). </strike>

@@

```{r umap-global-trends, fig.cap="Placeholder for figure 1", results=FALSE, eval=FALSE}
set.seed(21)
plot(x=rnorm(1000), y=rnorm(1000))
```

>>>>

---
title: "A molecular taxonomy of tumors independent of tissue-of-origin"
author: "Peter T. Nguyen$^{1}$, Simon G. Coetzee$^{1}$, Daniel L. Lakeland$^{2}$, and Dennis J. Hazelett$^{1,3}$"
output: 
    bookdown::word_document2:
    number_sections: FALSE
    toc: FALSE
bibliography: references.bib
---

@@

```{r echo=FALSE, eval=FALSE}
  bookdown::pdf_book:
    number_sections: FALSE
    toc: FALSE
    keep_tex: yes

  bookdown::html_document2:
    fig_captions: TRUE
    number_sections: FALSE
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
    code_folding: hide
    theme: flatly
    smart: true

  bookdown::word_document2:
    number_sections: FALSE
    toc: FALSE
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, dpi=600)
suppressPackageStartupMessages(c(
  library(rstan),
  library(foreach),
  library(sqldf),
  library(tidyverse),
  library(ggthemes),
  library(data.table),
  library(biomaRt),
  library(dplyr),
  library(GenomicRanges),
  library(SummarizedExperiment),
  library(reactome.db),
  library(plyr),
  library(cluster),
  library(pheatmap),
  library(clusterProfiler),
  library(org.Hs.eg.db),
  library(dbscan),
  library(class),
  library(TCGAbiolinks),
  library(patchwork),
  library(ggplotify),
  library(ComplexHeatmap),
  library(knitr),
  library(kableExtra),
  library(UpSetR),
  library(cowplot),
  library(doRNG),
  library(uwot),
  library(janitor),
  library(FactoMineR),
  library(seriation),
  library(rtracklayer),
  library(poibin),
  library(flextable)
))
```

@@

```{r child='../code/figure2.Rmd'}

```

@@

<strike>
### Gene-Pathway redundancy contributes to cluster identity

@@

@@

Third, we checked the mutation classes using MAFtools [@maftools] and found the overwhelming
majority (x out of y) are missense _vs._ various other deleterious classes (see
**figure 3D**), suggesting that, similar to other oncogenic driver classes such
as KRAS and PIK3C, proteasomal subunit mutations may result in specific
dysregulation of the proteasomal complex activities that target cancer-promoting
pathways. Taken together, our data suggest that mutations in proteasomal
subunits are more widespread in cancer than previously appreciated and may
represent a targetable category of tumors.
</strike>

```{r child = '../code/figure4.Rmd'}

```

@@

As discussed previously, prior efforts to extract common signatures from
pan-cancer datasets met with difficulty in distinguishing tumor samples from
their tissue-specific -omics data signatures. Given our pathway-disruption based
classification, this raises the question, are tumor phenotypes driven entirely
by common driver genes, or by "silent" tissue-specific effectors ( _i.e._ too
few samples to detect above statistical significance thresholds), or a
combination of both? To answer this question, we compared each cluster relative
to the background to find deferentially mutated genes. We ranked the odd ratios 
and selected the top ten enriched and depleted genes (pvalue < 0.01) for each 
cluster (**Figure 4**; odds ratios plot). We find that the clusters possess 
unique gene sets. Cluster 2 is enriched for genes responsible for protein
homeostasis. Cluster 4 is enriched for genes responsible for nucleosome 
structure. Clusters 7 and 8, which shared multiple enrichment in signaling
pathways, are largely driven by mutations in _PI3K_ and its homologs and _Ras_ 
genes, respectively (compare _PIK3CA_ and _KRAS_ panels of **Supplementary figure S5**). 
Interestingly, cluster 9 is enriched for both _PIK3CA_ and _KRAS_ which may
suggest tumors in this cluster derived from either clusters 7 or 8. Next, we 
investigated the proportion of samples per cancer type for the significant genes 
within a cluster (**Figure 4**; heatmap). Surprisingly, cluster-specific tumors 
were not predominated by one or more highly mutated genes across all cancers. 
Instead, some cancer-specific tumors in these classes had mutation profiles that
differ from one another (_e.g._ In cluster 2, LIHC was enriched for _PSME4_, KIRC
was enriched for _UBC_, and PAAD was enriched for _PSMA1_).

@@

Therefore, we calculated cluster-specific enrichment in metastatic tumors and
found a total of 27 enriched pathways across all clusters (**table 1**). A number of
enrichments were found in multiple clusters represented by pathways that were
already shown to be enriched in a neighboring cluster. For example,
“Signaling by PTK6” is enriched in non-metastatic samples of cluster 8 (see 
**supplementary figure S4**) but not in 7 and 9. This pathway is enriched in
metastatic tumors of clusters 7 and 9 (p < 10-3, **table 1**). This is also true of 
“Erythropoietin activates RAS”, which is enriched in cluster 8 (**supplementary figure
S4**) and also in metastatic tumors of clusters 7 and 9. <strike>Cluster 1, one of two clusters 
for which we could not identify any distinguishing pathways (cluster 8 being the other), 
had enrichment in two pathways that were otherwise specific to non-metastatic tumors of other
clusters. These include “Cellular response to hypoxia” (enriched in clusters 6 &
7) and “TNFR2 non-canonical NF-kB pathway” (cluster 6 & 7).</strike> Cluster 4 metastatic
samples was enriched for “Fc epsilon receptor (FCERI) signaling”, a key neutrophil pathways, 
which are specific to clusters 2, 7 and 8 non-metastatic tumors. Thus, pathway lesions 
specific to metastases tend toward complementary mutations that are otherwise specific to
non-metastatic tumors of other clusters.

```{r child='../code/table2.Rmd'}
```
<mark>Statement does not apply anymore. The highest # of enriched path is 3. Not many are in 
metastasis. Only 2. “Fc epsilon receptor (FCERI) signaling” and "Signaling by SCF-KIT" </mark>
<strike>There were also a number of pathways that were enriched with lower specificity
(among 4 or more clusters) which were found to be enriched in metastases. We
identified 18 such pathways of which 12 had significantly higher mutation rates
in cluster-specific metastases, including “G1/S DNA Damage Checkpoints”,
“Deubiquitination”, “Sumoylation”, “M Phase”, “RHO GTPase Effectors”, “TCF
dependent signaling in response to WNT”, “Class I MHC mediated antigen
processing & presentation”, “Chaperonin-mediated protein folding”, “Intrinsic
Pathway for Apoptosis”, “Cellular Senescence”, “DNA Double-Strand Break
Response”, and “Pre-NOTCH Expression and Processing” (**Table 1**). </strike>

@@
